---
title: "class11"
author: "Muhammad Tariq"
date: "2025-05-06"
output: html_document
---
Install packages that were installed in Class10
```{r}
install.packages("bio3d")
install.packages("NGLVieweR")
install.packages(Rtools)
install.packages("pak")
pak::pak("bioboot/bio3dview")

install.packages("BiocManager")
BiocManager::install("msa")
```

```{r}
writeLines('PATH="${RTOOLS40_HOME}\\usr\\bin;${PATH}"', con = "~/.Renviron")

pkgbuild::check_build_tools(debug = TRUE)
install.packages("pkgbuild")

```

Load up the pakages we will need for analysis of protein structure sets.

```{r}
library(bio3d)
```

We will analyze the ADK starting with a singe ADK 
code:"1ake_A"

```{r}
id<- "1ake_A"
aa<- get.seq(id)
aa
```
Now we can search the PDB database to find all related enteries.

```{r}
blast <- blast.pdb(aa)
```

```{r}
attributes(blast)
```

Make a little summary figure of these results
```{r}
hits <- plot(blast)
```

Our "top hits" i.e. the most similar enteries in the database are:

```{r}
hits$pdb.id
```

```{r}
# Download releated PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

Align and supperpose these structures

```{r}
# Align releated PDBs
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```

Side-note:
```{r}
library(bio3dview)

view.pdbs(pdbs)
```

This is better but still difficult to see what is similar and diffeent in all these structures or indeed learn much about how this family works.

Lets try PCA:

```{r}
pc <- pca(pdbs)
plot(pc)
```

```{r}
plot(pc, pc.axes=1:2)
```

```{r}
view.pca(pc)
```
write a PDB "trajectory" for mol-star

```{r}
mktrj(pc, file="pca_results.pdb")
```

