---
title: "Class10 strutural"
author: "Muhammad Tariq"
format: pdf
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(bio3d)
library(NGLVieweR)
library(bio3dview)
library(ggplot2)
library(ggrepel)
```

```{r}
library(bio3d)
pdb <- read.pdb("1hsg")
```


## Q1: Percentage solved by X-Ray and Electron Microscopy
```{r}
total_structures <- 203169 + 13047 + 14528 + 4687 + 213 + 22
xray_total <- 172081 + 10027 + 8849 + 2947 + 170 + 11
em_total <- 18135 + 2974 + 5386 + 185 + 10 + 0
xray_percent <- round((xray_total / total_structures) * 100, 2)
em_percent <- round((em_total / total_structures) * 100, 2)
list(XRay = xray_percent, EM = em_percent)
```

## Q2: Proportion of protein-only structures
```{r}
protein_only <- 203169
protein_proportion <- round((protein_only / total_structures) * 100, 2)
protein_proportion
```

## Q3: Number of HIV-1 protease structures
**Answer**: Use the [RCSB search for HIV-1 protease](https://www.rcsb.org/). At the time of writing, there are ~395 structures.

## Q4: Why do water molecules appear as single atoms?
**Answer**: X-ray crystallography typically resolves only the oxygen atom of water, hence only one atom per water molecule appears.

## Q5: Conserved water molecule residue number?
**Answer**: The conserved water molecule in 1HSG is residue number HOH 301 or similar (check Mol* under Ligands > Water > Hover labels).

## Q6: Save a figure showing both chains, ligand, ASP25 and critical water
**Answer**: Done via Mol*, screenshot saved manually and added to report.

## Q7: Binding Site Discussion
**Answer**: Indinavir and similar inhibitors enter the binding pocket via a flap opening mechanism in the protease dimer. Structural plasticity allows entry.

## Q8: Number of amino acid residues in 1HSG
```{r}
pdb <- read.pdb("1hsg")
length(unique(pdb$atom$resno[pdb$atom$type == "ATOM"]))
```

## Q9: Name a non-protein residue
```{r}
unique(pdb$atom$resid[!(pdb$atom$resid %in% c("ALA", "VAL", "ILE", "GLY", "GLU", "ASP", "TYR", "TRP", "PHE", "ASN", "GLN", "CYS", "HIS", "LYS", "ARG", "SER", "THR", "MET", "PRO"))])
```

## Q10: Which package is from Bioconductor?
**Answer**: `msa`

## Q11: Which package is from GitHub?
**Answer**: `bio3dview` (from `bioboot/bio3dview`)

## Q12: Can `pak` install from GitHub?
**Answer**: TRUE

## Q13: Length of ADK sequence (1AKE_A)
```{r}
aa <- get.seq("1ake_A")
length(aa$ali)
```

## Q14: Normal Mode Analysis observation
```{r}
hits <- list(pdb.id = c('1AKE_A','6S36_A','6RZE_A','3HPR_A','1E4V_A','5EJE_A','1E4Y_A','3X2S_A','6HAP_A','6HAM_A','4K46_A','3GMT_A','4PZL_A'))
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
rd <- rmsd(pdbs)
hc.rd <- hclust(dist(rd))
grps.rd <- cutree(hc.rd, k=3)
modes <- nma(pdbs)
plot(modes, pdbs, col=grps.rd)
```
**Answer**: The black and colored lines indicate variance in flexibility. They differ most in flexible loop regions—these regions correspond to conformational transitions important for Adk function.

## OPTIONAL: View PCA in interactive widget
```{r}
pc.xray <- pca(pdbs)
view.pca(pc.xray)
```

## OPTIONAL: Export trajectory and load into Mol*
```{r}
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
```
**Instructions**: Open `pc_1.pdb` in [Mol*](https://molstar.org/viewer/) and click the “play” icon to view animated PC1 motion.

## OPTIONAL: ggplot-based PCA Plot
```{r}
ids <- basename.pdb(pdbs$id)
df <- data.frame(PC1=pc.xray$z[,1], 
                 PC2=pc.xray$z[,2], 
                 col=as.factor(grps.rd),
                 ids=ids)

p <- ggplot(df) + 
  aes(PC1, PC2, col=col, label=ids) +
  geom_point(size=2) +
  geom_text_repel(max.overlaps = 20) +
  theme(legend.position = "none")
p
```
