---
title: "class12"
author: "Muhammad Tariq"
date: "2025-05-14"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Section 1. Proportion og G/G in a population

Download a CSV file from ENSEMBLE
https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:42324335-42325335;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel >

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```
 
```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100

```
 
Now lets look at a different opulation I picked the GBR.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

Find portion of G/G
```{r}
round(table(gbr$Genotype..forward.strand.) /nrow(gbr) * 100, 2)
```

This varient that is associated with childhood asthma is more frequent in the GBR population than the MKL population.

Lets now dig into this further.


##Section 4: Population Scale Analysis [HOMEWORK]
One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale.

How many samples do we have?
```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)

```

```{r}
library(ggplot2)

```
Lets make a boxplot
```{r}
ggplot(expr) +aes(geno, exp, fill=geno) +
  geom_boxplot(notch = TRUE)
```

