---
title: 'Class 8: Breast Cancer mini project'
author: "Muhammad Tariq"
date: "2025-04-24"
output:
  pdf_document: default
  html_document: default
---

## Background

Unsupervised learning, supervised learning and reinforcement learning; Focus on unsupervised learning, K-means clustering, Hierarchical clustering, Dimensionality reduction, visualization and analysis, Principal Component Analysis (PCA) Practical considerations and best practices for the analysis of high dimensional datasets.


##Data Import

Our data come from the U. of Wisconsin Medical Center

```{r}
wisc.df <-read.csv("WisconsinCancer.csv", row.names=1)
```
> Q1. How many patients /samples are in the dataset?

```{r}
nrow(wisc.df)
```


> Q2. How many of the observations have a malignant diagnosis? 

```{r}
table(wisc.df$diagnosis)
```
```{r}
sum(wisc.df$diagnosis== "M")
```


> Q3.  Q3. How many variables/features in the data are suffixed with _mean?

```{r}
colnames(wisc.df)
```
```{r}
length( grep("mean", colnames(wisc.df), value = T))
```

There is a diagnosis column that is the clinician consense that I want to exclude from any further analysis. We will come back later and compare our results to this diagnosis. 

```{r}
diagnosis <- as.factor(wisc.df$diagnosis)
head(diagnosis)
```
Now we can remove it from the 'wisc.df'

```{r}
wisc.data <-wisc.df[,-1]
```

## Clustering 

Lets try a hclust()
```{r}
hc <-hclust(dist(wisc.data))
plot(hc)
```

We can extract clusters from this rather poor dendrogram/tree with the 'cutree()'

```{r}
grps <-cutree(hc, k=2)
```

How many individuals in each cluster?

```{r}
table(grps)
```
```{r}
table(diagnosis)
```
We can generate a cross-table that compares our cluster 'grps' vector with out 'diagnosis'
vectore values

```{r}
table(diagnosis, grps)
```
## Principal Component Analysis
prcomp(x,scale=F,center=F)
### The importance of PCA Data
The main function for PCA in base R is 'prcomp()' it has a defualt input parameter of 'scale=FALSE'.


```{r}
#prcomp()
head(mtcars)
```

We could do a PCA of this data as is and it could be mis-leading....

```{r}
pc <- prcomp(mtcars)
biplot(pc)
```

Lets look at the mean values of each column and their standard deviation. 

```{r}
colMeans(mtcars)
```
```{r}
apply(mtcars, 2, sd)
```
We can "scale" this data before PCA to get a much better representation and anlysis of all the columns.

```{r}
mtscale <- scale(mtcars)
```

```{r}
round(colMeans(mtscale))
```
```{r}
apply(mtscale,2,sd)
```


```{r}
pc.scale <-prcomp(mtscale)
```

We can look at the two main results figures from PCA - the 'PCplot" (a.k.a. score plot, orientation plot, or pc1 vs pc2 plot). The "loadings plot" how the original variables contribute to the new PCS.

A loadings plot of the unscalled PCA results
```{r}
library(ggplot2)

ggplot(pc$rotation) +
  aes(PC1, rownames(pc$rotation)) +
  geom_col()
```

Loadings
```{r}
ggplot(pc.scale$rotation) +
  aes(PC1, rownames(pc$rotation)) +
  geom_col()
```
PC plot of scaled PCA results
```{r}
library(ggrepel)

ggplot(pc.scale$x) +
  aes(PC1,PC2, label=rownames(pc.scale$x)) +
  geom_point() +
  geom_text_repel()
```






> **Key point**: In general we will set 'scale=True' when we do PCA. This is not the defualt but probably should be....

We can check the SD and mean of the different columns in 'wisc.data' to see if we need to sclae - hint; we do!

### PCA of wisc.data

```{r}
wisc.pr <- prcomp(wisc.data, scale= TRUE)
```

To see how well PCA is doing here interms capturing the variance (or spread)  in the data we can use the 'summary()' function.


```{r}
summary(wisc.pr)
```
> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

Lets make the main PC1 VS PC2

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC2,col=diagnosis) +
        geom_point() +
  xlab("PC1 (44.3%))") +
  ylab("PC2 (19%)")
```
> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

PC1, PC2 and PC3 is required for at least 70%

>Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

PC1, PC2, PC3, PC4, PC5, PC6, PC7 and PC8. 

>Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

What stands out is that thewre is a clear significance in the causation on the casncer being benign and malignant depending on the data, once creating the PCA axis it will highlight it more to us.

>Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC2,col=diagnosis) +
        geom_point() +
  xlab("PC1 (44.3%))") +
  ylab("PC3 (9.4%)")
```

>Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation["concave.points_mean", 1]
```

>Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

```{r}
summary(wisc.pr)

```
It would include PC30-PC9 which includes rest of 20%

## 5. Combining methods

### Clustering on PCA results

```{r}
wisc.pr.hclust <- hclust( dist(wisc.pr$x[,1:2]), method="ward.D2")
plot(wisc.pr.hclust)
```

We can "cut" this tree to yield out cluster (groups):
```{r}
pc.grps <-cutree(wisc.pr.hclust, k=2)
table(pc.grps)
```
How do my cluster grps compare to the expert diagnosis

```{r}
table(diagnosis,pc.grps)
```

```{r}
table(diagnosis)
```
>Q15. How well does the newly created model with four clusters separate out the two diagnoses?

```{r}
# Run k-means with 4 clusters on PCA-reduced data
set.seed(123)  # For reproducibility
wisc.km.4 <- kmeans(wisc.pr$x[, 1:3], centers = 4)

# Compare clusters to actual diagnoses
table(wisc.km.4$cluster, diagnosis)
```
It did it really well, had a shown a great differences in our patients diagnosis through the family tree and if needed we could further make the groups even more specific.

>Q16. How well do the k-means and hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.

They did really badly. We do much better after PCA - the new PCA variables (what we call a basis set) give us much better separation of M and B


##7. Prediction

We can use out PCA model for the analysis of new "unseen" data. In this case from U. Mich.
```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```
> Q18. Which of these new patients should we prioritize for follow up based on your results?

```{r}
head((npc))

prioritize <- npc[, "PC1"] < -5  
# adjust threshold based on earlier plots

npc[prioritize, ]
```
Patients with strongly negative PC1 scores — for example, those with PC1 < -5 — should be prioritized for follow-up. These patients are projected close to where malignant cases cluster in the PCA analysis and may have a higher risk of malignancy.

