---
title: "Week 4 R functions"
author: "Muhammad Tariq"
date: "2025-04-24"
output: pdf_document
---

This week we are introducing **R functions** and how to write our own functions. 
Questions to answer:

> Q1. Write a function grade() to determine an overall grade from a vector of student homework
assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an
NA value) this can be used as a score to be potentially dropped. Your final function should be
adquately explained with code comments and be able to work on an example class gradebook
such as this one in CSV format: “https://tinyurl.com/gradeinput” [3pts]

```{r}
plot(1:10)
```


```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)

student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)

student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```
 
Follow the guidelines from class

- Write a working snipet of code that solves a simple problem

```{r}
#Straight forward mean()
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)

mean(student1)
```
But.... We need to drop the lowest score. First we identify the lowest score. 

```{r}
# Which element of the vector is the lowest?
which.min(student1)
```
What I want is to now drop (i.e. exclude) this lowest score from my mean() calculation. 

```{r}
# This will return everything but the 8th element of vector
student1[-8]
```
Now we can use the answer from which.min() to return all other elements of the vector

```{r}
# This is out first working snippet
mean(student1[-which.min(student1)])
```

What about the other example students? Will this work for them?

We could try using na.rm=TRUE argument for mean but this is pants! Not a good enough approach i.e unfair. 
```{r}
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
mean(student2,na.rm=TRUE)

student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
mean(student3,na.rm=TRUE)

```

Another approach is to mask (i.e replace) all NA values with zero

First we need to find the NA elements of the vector. How do we find the NA elements?
```{r}
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
x <- student2

is.na(x)
which( is.na(x))
```

Now we have identified the NA elements we want to "mask" them. Replace them with zero?

```{r}
#This does not quite get us there
mean(x[-which(is.na(x))])
```
Instead we will make the NA elements zero

```{r}
# Cool this is useful
x[is.na(x)] <- 0
x
mean(x)
```

Recall we should drop the lowest score now...

```{r}
x[is.na(x)] <- 0
mean(x[-which.min(x)])

```
Now we are essentially there with our working snippet
```{r}
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
x <- student3
x[is.na(x)] <- 0
mean(x[-which.min(x)])
```

## Now we make our function

Take the snippet and turn into a function every function has 3 parts

-A name, in our 'grade()'
-Input arguments, a vector of student scores
-The body i.e. our working snippet code

Using R studio I will select 'Code > Exract Functon'
```{r}
grade <- function(x) {
  x[is.na(x)] <- 0
  mean(x[-which.min(x)])
}
```

```{r}
grade(student1)
grade(student2)
grade(student3)

```

This looks great ! we no need to add comments to explain this to our future selves and others who want to use this function.

```{r}
#' Calculate average score for a vector of student scores dropping the lowest score. Missing values will be treated as zero.
#'
#' @param x A numeric vector of homework scores
#'
#' @returns Average score
#' @export
#'
#' @examples
#' student<-c(100,NA,90,97)
#' grade(student)
#' 
grade <- function(x) {
  #Mask NA with zero 
  #Treat missing values as zero
  x[is.na(x)] <- 0
  #Exclude lowest score for mean
  mean(x[-which.min(x)])
}
```
Now finally we can use our function on our "real" whole class data from this CSV format: “https://tinyurl.com/gradeinput”

```{r}
url <- "https://tinyurl.com/gradeinput"
gradebook <- read.csv(url, row.names = 1)
```

```{r}
apply(gradebook, 1, grade)
```


>Q2.Using your grade() function and the supplied gradebook, Who is the top scoring student
overall in the gradebook?

To answer this we run the apply() function and save the results. 


```{r}
results <-apply(gradebook, 1, grade)
sort(results, decreasing = TRUE)
```

```{r}
which.max(results)
```
>Q3. From your analysis of the gradebook, which homework was toughest on students (i.e. obtained
the lowest scores overall?

```{r}
gradebook
```

```{r}
ave.scores <-apply(gradebook, 2, mean, na.rm=TRUE)
ave.scores
which.min(ave.scores)
```

```{r}
med.scores <- apply(gradebook, 2, median, na.rm=TRUE)
med.scores
which.min(med.scores)
```

```{r}
boxplot(gradebook)
```
> Q4. Optional Extension: From your analysis of the gradebook, which homework was most
predictive of overall score (i.e. highest correlation with average grade score)?

Are the final results (i.e. average score for each student) correlated with results (i.e. scores) for individual homeworks - the gradebook columns

```{r}
masked.gradebook <- gradebook
masked.gradebook[ is.na(masked.gradebook)] <- 0
masked.gradebook
```

And look at corelation
```{r}
cor(results, masked.gradebook$hw5)
```

```{r}
apply(masked.gradebook, 2, cor, x=results)
```



> Q5. Make sure you save your Quarto document and can click the “Render” (or Rmarkdown”Knit”) button to generate a PDF foramt report without errors. Finally, submit your PDF
to gradescope.

Knit the document to make a PDF 